---
title: "Multispecies optimisation: methods"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    df_print: kable
    fig_caption: true
    latex_engine: xelatex
mainfont: Calibri 
fontsize: 11pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing  
---
```{r include=FALSE}
#set global options for code chunks
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

#required packages
#tinytex::install_tinytex()
#tinytex::reinstall_tinytex()
library(tinytex)
source("../scripts/optCalcFunctions.R") 

```



# Results

 + Preliminary graphs showing how the optimal alpha for species A and B varies with either; a change in density of species A, or a change in the effective strip width of species A (all other parameters are kept constant).
 

```{r}

#Visualising the driver of the optimal approach: at what point (abundance?)
# for species B do you measure everything?

# cost per unit length of transect (Cl) between 1 second and 10 seconds per m
# cost to measure distance to 1 target between 15 seconds and 1 minute

ratClCm <- expand.grid("Cl" = seq(1/60, 10/60, by = 3/60),
                       "Cm" = seq(0.25, 1, by = 0.25))
#this isn't good - think through this in a more systematic way...  
#5 different cost ratios, based on five different circumstances:
# low cost of walking, moderate cost of measuring (detections in quick succession, but not so frequent that a wide strip is stupid, so measuring could be a decent distance)
# super low cost of walking (like, detections every few cm), low cost of measuring (this may help to clarify what influence the cost ratio has, and will represent the very high density small strip width scenarios - we could also try to work in some 'aggregation' index to account for detections close to each other taking less time to measure?)
# high cost of walking, moderate cost of measuring 
# moderate cost of walking, moderate cost of measuring
# moderate cost of walking, high cost of measuring


ratClCm$costMult <- ratClCm$Cl/ratClCm$Cm

popPars <- expand.grid("D" = seq(0.005, 0.5, by = 0.05),
                       "sigm" = seq(2, 20, by = 2))
popPars$mu <- sqrt(pi*(popPars$sigm^2)/2)

midMinMax <- expand.grid("costMult" = median(ratClCm$costMult),
                         "D" = c(min(popPars$D), median(popPars$D), max(popPars$D)),
                         "mu" = c(min(popPars$mu), median(popPars$mu), max(popPars$mu)))

#plot 1-4: species B is low density, mid detectability
Da <- seq(0.005, 0.5, by = 0.005)
Da1 <- 0.355
muA <- seq(0.25, 25, by = 0.25)
muA1 <- 25.066283
Db1 <- 0.005
muB1 <- 12.533141

plotY1 <- altoptAoptB(midMinMax$costMult[2], muA = muA, muB = muB1, Da = Da1, Db = Db1)
plotY2 <- altoptAoptB(midMinMax$costMult[2], muA = muA1, muB = muB1, Da = Da, Db = Db1)

```


```{r, fig.cap= "Optimal alpha over a range of values for mu of species A"}

 par(mfrow = c(1,2), mar = c(5,4,2,2), oma = c(0,0,2,0), pty = "s")
 plot(x = muA, y = plotY1$alphA, xlim = c(0, max(muA)), ylim = c(0, 1), lty = 1, 
      ylab = "opt alpha A", type = "l", lwd = 2, cex.axis = 0.8)
 plot(x = muA, y = plotY1$alphB, xlim = c(0, max(muA)), ylim = c(0, 1), type = "l", 
      ylab = "opt alpha B", lwd = 2, cex.axis = 0.8)
 mtext(paste0("Da = ", Da1, "; Db = ", Db1, "; muB = ", muB1), outer = TRUE, side = 3, line = 0)

```


```{r, fig.cap="Optimal alpha over a range of values for density of species A"}

par(mfrow = c(1,2), mar = c(5,4,2,2), oma = c(0,0,2,0), pty = "s")
plot(x = Da, y = plotY2$alphA, xlim = c(0, max(Da)), ylim = c(0, 1), type = "l", 
     ylab = "opt alpha A", lwd = 2, cex.axis = 0.8)
plot(x = Da, y = plotY2$alphB, xlim = c(0, max(Da)), ylim = c(0, 1), type = "l", 
     ylab = "opt alpha B", lwd = 2, cex.axis = 0.8)
mtext(paste0("muA = ", muA1, "; Db = ", Db1, "; muB = ", muB1), outer = TRUE, side = 3, line = 0)
```

#thinking about the 3 or 4 different lines: characteristics of species A and B

#species A higher density and detectability (no need to do the reverse?) - 2
#species A and B are similar - 3
#species A is similar density but more detectable, and/or vice versa - 4
  #no: this is covered by the previous graphs
#the ratio muB/muA (can look at the above scenarios to see how much it explains)


```{r, fig.cap="Varying mu of species A over three scenarios"}
D_vect <- seq(0.005, 0.5, by = 0.005)
mu_vect <- seq(0.25, 25, by = 0.25)

# 2: species A higher density and detectability (max vs mid, A vs B)
Da2 <- max(D_vect)
muA2 <- max(mu_vect)
Db2 <- median(D_vect)
muB2 <- median(mu_vect)
plotY2_mu <- altoptAoptB(median(ratClCm$costMult), muA = mu_vect, muB = muB2, Da = Da2, Db = Db2)
plotY2_D <- altoptAoptB(median(ratClCm$costMult), muA = muA2, muB = muB2, Da = D_vect, Db = Db2)

# 3: species A and B are similar (but not identical)
Da3 <- median(D_vect)-0.05
muA3 <- median(mu_vect)-0.5
Db3 <- median(D_vect)+0.05
muB3 <- median(mu_vect)+0.5
plotY3_mu <- altoptAoptB(median(ratClCm$costMult), muA = mu_vect, muB = muB3, Da = Da3, Db = Db3)
plotY3_D <- altoptAoptB(median(ratClCm$costMult), muA = muA3, muB = muB3, Da = D_vect, Db = Db3)

#4: species A higher (D and mu) but less difference than min/max
Da4 <- min(D_vect)+0.15
muA4 <- min(mu_vect)+5
Db4 <- max(D_vect)-0.15
muB4 <- max(mu_vect)-5
plotY4_mu <- altoptAoptB(median(ratClCm$costMult), muA = mu_vect, muB = muB4, Da = Da4, Db = Db4)
plotY4_D <- altoptAoptB(median(ratClCm$costMult), muA = muA4, muB = muB4, Da = D_vect, Db = Db4)

speciesPars <- data.frame("muA" = c(muA2, muA3, muA4), "Da" = c(Da2, Da3, Da4), 
                       "muB" = c(muB2, muB3, muB4), "Db" = c(Db2, Db3, Db4), 
                       row.names = c("solid", "dashed", "dotted"))

#write.table(speciesPars, "../graphs/MultiSpMultiLinePars.txt")
knitr::kable(speciesPars, caption = "Species A and B parameters")

alphABplots <- function(vect, Y1, Y2, Y3, vectname){
  par(mfrow = c(1,2), mar = c(5,4,2,2), oma = c(0,0,2,0), pty = "s")
  
  plot(x = vect, y = Y1[,1], ylim = c(0,1), xlim = c(0,max(vect)),
       ylab = "opt alpha A", xlab = vectname,  
       type = "l", lwd = 2, cex.axis = 0.8)
  lines(x = vect, y = Y2[,1], lty = 2, lwd = 2)
  lines(x = vect, y = Y3[,1], lty = 3, lwd = 2)
  legend(legend = c("high A mid B", "A and B similar", "low A high B but closer"),
         lty= c(1,2,3), x = "top", cex = 0.7)
  
  plot(x = vect, y = Y1[,2], ylim = c(0,1), xlim = c(0,max(vect)),
       ylab = "opt alpha B", xlab = vectname,  
       type = "l", lwd = 2, cex.axis = 0.8)
  lines(x = vect, y = Y2[,2], lty = 2, lwd = 2)
  lines(x = vect, y = Y3[,2], lty = 3, lwd = 2)
  legend(legend = c("high A mid B", "A and B similar", "low A high B but closer"),
         lty= c(1,2,3), x = "top", cex = 0.7)
  
}

#plots 1 and 2
#png("graphs/MultiSpMultiLineMu.png", width = 1000, height = 500)
alphABplots(vect = mu_vect, Y1 = plotY2_mu, Y2 = plotY3_mu, Y3 = plotY4_mu, vectname = "muA")
#dev.off()

```


```{r, fig.cap="Varying density of species A over three scenarios"}
#png("graphs/MultiSpMultiLineD.png", width = 1000, height = 500)
alphABplots(vect = D_vect, Y1 = plotY2_D, Y2 = plotY3_D, Y3 = plotY4_D, vectname = "Da")
#dev.off()
```

Want to plot the objective function: plot varDA + varDB (x) against alpha A, keeping alpha B constant, and vice versa.